Supplementary Information

### 2021-02-09 doc/manuscript/MEE_manuscript_SI.Rmd
wrote Appendix S2

*This is still rough/ mostly notes.*

In most cases, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings [@applequist_simple_1958; @duncan_evaluation_1989], $DBH$ was reconstructed from the inside out.

For each core, $DBH$ can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $RW$ from the inside out,--only when core hit pith or distance to pith can be reliably estimated). 
We generally gave precedence to the outside-in approach.
Specifically, when $DBH$ was taken at the time of coring,  
At some of our sites where DBH was not taken at the time of coring (*SCBI*,), DBH measurements taken before or slightly after the time of coring could be used.
(see [issue #19 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/19))
If before, ... 
If after...
For all outside-in reconstructions, if a negative $DBH$ was predicted...

When there were more than one cores for a tree, the $DBH$ reconstructions from each core were averaged to produce a single estimate of the tree's $DBH$ through time. 
When the start or end dates of the records from the cores differed, we extrapolated growth of the shorter core to match the years covered by the longer core. 
Specifically, to fill in years at the more recent end, we assumed that the average growth rate of the ten years prior to the missing records applied to the missing years. 
To fill in years at the beginning of the tree's lifespan, we likewise assumed that the ten years adjacent to the missing record applied to the missing years; however, if this yielded a negative $DBH$ estimate for the earliest year in the reconstruction, we divided the existing minimum $DBH$ by number of years missing and applied that value to each year.
We note that these reconstructed growth records were used only for the reconstruction of $DBH$ and were not included as response variables in any of our analyses.

In either case we need bark thickness--ideally allometries describing the relationship between DBH and bark thickness (Table S4). 
This is especially critical for thick-barked species.
When bark thickness data were available, we generated allometries ([issue #8 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/8))... 
lognormal model with intercept forced to zero: `lm(bark_depth.mm ~ -1 + log(dbh_no_bark.cm+1):bark_species, data = bark)`.
When bark thickness data were not available, we used published bark allometries from other sources (Table S4)

>>>>

*(documented in [GitHub issue #19 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/19))*

Tree diameter at breast height (or height of coring), $DBH$, was reconstructed over the time span for which a core record existed.
$DBH$ was reconstructed either from the outside in or from the inside out.
We gave precedence to the outside-in approach; that is, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings [@applequist_simple_1958; @duncan_evaluation_1989], $DBH$ was reconstructed from the inside out.

The outside-in approach to $DBH$ reconstruction started with a recent $DBH$ measurement, $DBH_m$, where reference measurement year $m$ is often but not necessarily at the time or coring.
$DBH$ for each year ($y$) of interest, $Y$, was  then estimated using the following equation:

$$DBH_Y  = DBH_{m} - 2*\left[r_{bark,m} - r_{bark,Y} + \sum_{y=Y}^{m} RW_{y} \right]$$
Here, $r_{bark}$ denotes bark thickness, which was estimated on a species-specific basis from allometries (Tables `r species_table`, `r bark_allometries`).
If $DBH_m$ was taken before coring, $RW$ values between $y=m$ and $y=Y$ were added instead of subtracted.
**If after...**

In some cases, negative values of $DBH_Y$ were predicted for the earliest years in the core record; that is, the sum of $RW$ from the earliest year in the core record to year $m$, plus the predicted change in $r_{bark}$, was greater than half of $DBH_m$. 
Such cases could occur due to a measurement or data error, or if the tree's growth was asymmetrical around the pith, as commonly occurs [@speer_fundamentals_2010]. 
In cases where the error was egregious -- i.e., where predicted $DBH_Y$ of the earliest year in the record divided by $DBH_m$ was < -0.5 -- we removed the tree from the analysis.
We assumed that smaller errors were attributable to asymmetrical growth, and re-distributed the error evenly across years so that $DBH_Y$ prior to the first year in the core was zero. 
This correction was applied only to $DBH_Y$, and not to any $RW$ records.

When a measurement of $DBH_m$ was not available, and when the core hit pith or the distance to pith could be reliably estimated, $DBH$ was reconstructed from the inside-out. 
Specifically, 
$DBH$ for each year ($y$) of interest, $Y$, was estimated using the following equation:

$$DBH_Y  = 2*\left[r_{bark,Y} + \sum_{y=0}^{Y} RW_{y} \right]$$
Again, $r_{bark}$ was estimated on a species-specific basis from allometries (Tables `r species_table`, `r bark_allometries`).

When there was more than one core for a tree, the $DBH$ reconstructions from each core were averaged to produce a single estimate of the tree's $DBH$ through time. 
When the start or end dates of the records from the cores differed, we extrapolated growth of the shorter core to match the years covered by the longer core. 
Specifically, to fill in years at the more recent end, we assumed that the average growth rate of the ten years prior to the missing records applied to the missing years. 
To fill in years at the beginning of the tree's lifespan, we likewise assumed that the ten years adjacent to the missing record applied to the missing years; however, if this yielded a negative $DBH$ estimate for the earliest year in the reconstruction, we divided the existing minimum $DBH$ by number of years missing and applied that value to each year.
We note again that these reconstructed growth records were used only for the reconstruction of $DBH$ and were not included as response variables in any of our analyses.

Regardless of the $DBH$ reconstruction approach (outside-in or inside-out), accurate reconstructions require accounting for $r_{bark}$, particularly for species that can attain very thick bark (e.g., *Pseudotsuga menziesii*, *Pinus ponderosa*).
For the majority of species, including all thick-barked species, we assigned allometries relating $r_{bark}$ to $DBH$, which were constructed for the species of interest or a related species (Tables `r species_table`, `r bark_allometries`).
When raw data on bark thickness data were available, we generated species-specific lognormal scaling relationships with the intercept forced to zero (Table `r bark_allometries`). 
When raw bark thickness data were not available, we applied published bark allometries from other sources (Table `r bark_allometries`).
